Percussive Sound Generation with Virtual Listeners and Modular Synthesizers

### 2020-08-20 templates_aimusic2020/paper.tex
Updates from Overleaf

\subsection{Motivation and Goals}
Digital recordings of novel, one-shot\footnote{A single hit on the drum that captures its capabilities} drum sounds are commonly used in  electronic music compositions. Yet unique drum sounds are not easy or cheap to find. By relying on recordings of \textit{real life} drum sounds, artitsts are limited by what instruments exist in the real world and whether or not high-quality, one-shot recordings can be accessed. Virtual sources of novel drum sounds can alleviate these material limitations.

@@

Numerous deep, neural network models have been proposed and utilized for the purpose of signal generation in recent years. WaveGans and WaveNet have been subject to significant improvements and experiments since their proposal~\cite{nsynth2017,yamamoto2020parallel,oord2017parallel}. Particularly relevant are the utilization of variational autoencoders (VAE's) for generation of percussive samples~\cite{aouameur2019neural} and generation of percussive sounds by decoding a small set of latent features~\cite{ramires2020neural}.
Automatic programming of virtual synthesizers has long been a topic of interest. In early 2000s, Interactive Genetic Algorithms (IGA's) were utilized for the generation of new sounds with various sound-engines~\cite{johnson1999exploring,dahlstedt2001creating}. More recent work by Yee-King et al.~\cite{yee2018automatic} used Long short-Term Memory (LSTM) models and genetic algorithms to find the exact parameters used to create a group of sounds. The sounds approximated were made by the same virtual synthesizer, not an external source; making the eventual replication certain even with random search. Since this work appears more focused on pads and textures rather than drums, feature matching appears to not be concerned with the envelope of the sounds but rather the frequency content within arbitrary time windows. Yet another recent work by Esling et al. used a large dataset of over 10,000 presets for a coMEMrcial VST synthesizer to learn a latent parameter space which can be sampled for creation of new audio~\cite{esling2019universal}. As stated before, our work explores the rapid approximation of percussion sounds with no previous knowledge about the sonic capabilities of our virtual synthesizer, exploring the actual parameter space rather than its latent representation.
\label{related_works}
\begin{center}
\begin{table}
\resizebox{\linewidth}{!}{\begin{tabular}{|c c c c|} 
\hline
Work & Feature extraction & Synthesis & Specilization & \hline
Oord 2017 & CNN & CNN &Speech& \hline
Yee-King 2018 & LSTM on Paremters & DSP & Synth Pads & \hline
Aouameur 2019 & Latent layer& Decoding of Latent Layers & Percussion & \hline
Ramires 2020 & Latent layer & FeedForward Network & Percussion & \hline
Yamamoto 2020 & GAN & GAN & Speech & \hline
\end{tabular}}
\caption{Quick reference of recent related works}
\end{table}
\label{table:recent_works}
\end{center}
\subsection{Our Approach}
To virtually create novel drum sounds, we need a tractable, generative source of audio; One that deterministically produces audio based on a set of instructions. Additionally, we need a method of evaluation to help us determine which sounds resemble drums and are worth keeping. To enable heuristic search, we would like to know what instructions caused our audio source to make the sounds we liked. Given these two components capable of \textit{generation} and \textit{evaluation} of sounds, we build a generative pipeline of audio as depicted in figure~\ref{fig:pipeline_outline}. The success of this pipeline is dependent on the characteristics of the components responsible for the generation and evaluation of the sounds. We call these two components the \emph{virtual synthesizer} and \emph{virtual ear} respectively.

@@

Sound is product of physical disturbances, causing vibrations in our mediums. Vibrations are perpetuated through air as part of an expanding, spherical front ~\cite{cook1999chap4}. A sound wave can be viewed as the result of a function which governs amplitude through time, where time and amplitude exist in continuous dimensions. Waves can be approximated via a series of samples, associating time steps to a discrete range of amplitude values.~\textit{Digital synthesis} of audio is the process of creating these discrete values.
Unlike the majority of the works mentioned in Section~\ref{related_works}, we do not employ probabilistic models for the synthesizer component. Our decision is based on the following factors:
\begin{itemize}
    \item \textit{Novelty and Creativity}: Given large sets of examples, generative models such as VAEs and GANS have proven capable of generating samples similar to the exemplars. In this work, we are not aiming for perfect imitations; Following Boden's definition of exploratory creativity as an emergent property of generative work within confined rule sets~\cite{boden2009computer}, we work within the limitations of any tractable sound source to create its  approximations of a given sound category. Within the music realm, prominent instance of exploratory creativity is the persistent aesthetics of 8-bit soundtracks, remaining popular long after the original constraints are lifted~\cite{collins2007loop}.
    \item \textit{Interpretability}: Neural networks are often described as black boxes~\cite{basheer2000artificial}. Their highly recursive structure makes modern explanation methods such as saliency maps unreliable~\cite{rudin2019stop}.  
    \item \textit{Speed of Rendering at high sampling rates}: Despite the utilization of powerful GPUs, the standard sampling rate in most audio generation work utilizing neural is under 24 khz \cite{yamamoto2020parallel,oord2017parallel,aouameur2019neural,ramires2020neural}. However, a significant number of untrained human ears can detect a change in quality of audio between sampling rates of 192 khz and the industry standard of 44.1 khz \cite{reiss2016meta}. In this work, we fix our sampling rate to the 48 khz standard. 
\end{itemize}

% \begin{figure*}[htbp]
% \label{fig_example_sine}
% \centering
% \includegraphics[width=0.3\linewidth,angle =-90 ]{images/periodic_function.png}
% \caption{A computer can simulate waveforms by utilizing periodic functions. Digital waveforms are discrete approximations of analogue waves~\cite{mitchell2009basicsynthChap5} }
% \end{figure*}

Evaluation of a periodic function such as sine or cosine is the simplest form of software audio generation~\cite{mitchell2009basicsynthChap5}. This generative system is called an \textit{oscillator}. The combination and modification of oscillators are the building blocks of digital signal processing (DSP). Most sounds from the natural environment, and its flora and fauna are more complex than the output of a single oscillator. However, with careful programming, DSP techniques can be used to replicate almost any sound, which is why they have powered coMEMrcial digital synthesizers for over half a century ~\cite{jenkins2019analog}. Another advantage to using DSP for sound generation is that the synthesizers we build using these functions are tractable: the output is determined by the input and reproducible. This makes the evaluation of a set of inputs (or parameters) to our synthesizers  simpler compared to the evaluation of synthesizers that utilize probabilistic models.

We opted for a set of classical DSP methods to build our synthesizer. For our project, we used the python based Pippi library for sound generation\footnote{https://github.com/luvsound/pippi}. This library uses a C back-end\footnote{https://github.com/PaulBatchelor/Soundpipe} and focuses on fast offline generation of audio signals. Our virtual synthesizer contains a set of one or more submodules. Each submodule is a self-contained noise making unit. Submodules have identical sets of parameters, but widely different outputs can be achieved depending on the values assigned. The set of parameters available to each submodule is highlighted in Table~\ref{table:submodule_params}. The sonic output of the virtual synthesizer is the normalized addition of the sonic output of its submodules. Our implementation of a synthesizer can have any number of submodules. The parameters that dictate the output signal of each submodule as well as the range of values each parameter can take are shown in table~\ref{table:submodule_params}. We call the number of submodules in each virtual synthesizer the \textit{stack size}. We call the sets of parameter values that characterize a synthesizer's submodules a \textit{program} (analogous to a preset for a VST).

@@

\begin{figure}[htbp!]
    \begin{center}
    \textbf{Data Overview}
    \makebox[\textwidth]{
    \includegraphics[width=0.95\linewidth]{images/venn_data.pdf}}
    \end{center}
    \caption{ 
    Scales are symbolic (true scales are not measurable and/or subjective).  $\mathcal{N}$ is the set of percussive sounds a synthesizer is capable of making. Our positive samples, $\mathcal{T^{+}}$, is a small fraction of a wide variety of percussive sounds that are conceivable. The size of $\mathcal{T^{-}}$ can be set to any number(we can generate as any number of random samples), but we prefer to keep it no larger than  $\mathcal{T^{+}}$. $\mathcal{H}$ is a series of sounds sent to the ear for classification. The ideal ear will detect  $\mathcal{H^{+}}$, the overlap of $\mathcal{H}$ and $\mathcal{N}$, with high precision and recall. 
    }
\label{fig:ven_data}
\end{figure}
\hypref{Decision.1} requires knowledge of what drums \textbf{do not} sound like, or knowledge of an infinitely large set, which cannot be fully represented via examples. An important consideration is that the source of sounds used in training the model (organic drum sounds) will be fundamentally different from the source of unlabeled sounds we wish to categorize (noise from a synthesizer). This issue is reflective of the OSR problem~\cite{geng2020recent,mundt2019open}. Figure~\ref{fig:ven_data} highlights a number of caveats with this approach. we denote $\mathcal{N}$ as the set of percussive sounds a synthesizer is capable of making. $\mathcal{T^{+}}$, our positive examples, is a set of sound extracted from material drums. $\mathcal{T^{-}}$, is a small subset of an infinite set which is meant to represent all noises which the synthesizer is capable of making and likely includes sounds in $\mathcal{N}$. During the hearing test, the virtual ear continually receives synthesizer noises for classification. The set of received noises is denoted as $\mathcal{H}$. The performance of the ear is measured by its precision and recall in finding the overlap between $\mathcal{H}$ and $\mathcal{N}$. It is important to ensure that the change in learning domains ---particularly with the case of positive examples of percussive sounds--- does not interfere with transformation of knowledge from the training of the ear to the hearing test. To maximize the transference of knowledge from one domain to another, we seek to learn from agnostic feature sets that capture fundamental characteristics of the data points.
\subsection{Feature Extraction}
To train our virtual ear, we require generalizable, domain agnostic features. Various works have demonstrated effective reconstruction of signals given their Short-time Fourier Transforms (STFT) ~\cite{nawab1983signal,griffin1984signal}. If the STFT of a signal can be used its reconstruction, perhaps it can be utilized as a source of fundamental features necessary for audible signal categorization. We defined 3 STFT transformation functions which we believe to capture important, unique attributes of percussive sounds (figure ~\ref{fig:stackspectrums}). Our transformations are GPU based, allowing their application in tandem to the training of the classifier or tuning of STFT transformation during hyper-paramter optimization (see Appendix~\ref{appendix:hyperparam}).
\subsection{Ear Types}
\label{classifiers}

>>>>

\subsection{Motivation and Goals}
Digital recordings of novel, one-shot\footnote{A single hit on the drum that captures its capabilities} drum sounds are commonly used in  electronic music compositions. Yet unique drum sounds are not easy or cheap to find. By relying on recordings of \textit{real life} drum sounds, artists are limited by what instruments exist in the real world and whether or not high-quality, one-shot recordings can be accessed. Virtual Studio Technology (VST)~\cite{tanev2013virtual} and other DSP based software have given artists unprecedented access to complex audio synthesizers. Utilization of virtual synthesizers for creation of novel drum sounds can alleviate these material limitations.

@@

Numerous deep, neural network models have been proposed and utilized for the purpose of signal generation in recent years. WaveGans and WaveNet have been subject to significant improvements and experiments since their proposal~\cite{nsynth2017,yamamoto2020parallel,oord2017parallel}. Particularly relevant are the utilization of variational autoencoders (VAE's) for generation of percussive samples~\cite{aouameur2019neural} and generation of percussive sounds by decoding a small set of latent features~\cite{ramires2020neural}. Automatic programming of virtual synthesizers has long been a topic of interest. In early 2000s, Interactive Genetic Algorithms (IGA's) were utilized for the generation of new sounds with various sound-engines~\cite{johnson1999exploring,dahlstedt2001creating}. More recent work by Yee-King et al.~\cite{yee2018automatic} used Long short-Term Memory (LSTM) models and genetic algorithms to find the exact parameters used to create a group of sounds. The sounds approximated were made by the same virtual synthesizer, not an external source; making the eventual replication certain even with random search. Since this work appears more focused on pads and textures rather than drums, feature matching appears to not be concerned with the envelope of the sounds but rather the frequency content within arbitrary time windows. Yet another recent work by Esling et al. used a large dataset of over 10,000 presets for a commercial VST synthesizer to learn a latent parameter space which can be sampled for creation of new audio~\cite{esling2019universal}.

\label{related_works}
\begin{table}[t!]
\begin{center}
\begin{tabular}{|c c c c|} 
\hline
Work & Feature extraction & Synthesis & Specilization \\ \hline
Oord 2017 & CNN & CNN & Speech \\  \hline
Yee-King 2018 & LSTM on Paremters & DSP & Synth Pads \\  \hline
Aouameur 2019 & Latent layer& Decoding of Latent Layers & Percussion \\  \hline
Ramires 2020 & Latent layer & FeedForward Network & Percussion \\  \hline
Yamamoto 2020 & GAN & GAN & Speech \\  \hline
\end{tabular}
\caption{Quick reference of recent related works}
\label{table:recent_works}
\end{center}
\end{table}
\subsection{Our Approach}
We explore the approximation of percussive sounds with no previous knowledge about the sonic capabilities of a virtual synthesizer. To virtually create novel drum sounds, we need a tractable, generative source of audio; One that deterministically produces audio based on a set of instructions. Additionally, we need a method of evaluation to help us determine which sounds resemble drums and are worth keeping. To enable heuristic search in future works, we would like to know what instructions caused our audio source to make the sounds we liked. Given components capable of \textit{generation} and \textit{evaluation} of sounds, we build a generative pipeline of audio as depicted in figure~\ref{fig:pipeline_outline}. The success of this pipeline is dependent on the characteristics of the components responsible for the generation and evaluation of the sounds. We call these two components the \emph{virtual synthesizer} and \emph{virtual ear} respectively.

@@

Sound is our experience of physical disturbances, causing vibrations in our mediums ~\cite{cook1999chap4}. A sound wave can be viewed as the result of a function which governs amplitude through time, where time and amplitude exist in continuous dimensions. Waves can be approximated via a series of samples, associating time steps to a discrete range of amplitude values. \textit{Digital synthesis} of audio is the process of creating these discrete values. Unlike the majority of the works mentioned in Section~\ref{related_works}, we do not employ probabilistic models for the synthesizer component. Our decision is based on the following factors:
\begin{itemize}
    \item \textit{Novelty and Creativity}: Given large sets of examples, generative models such as VAEs and GANS have proven capable of generating samples similar to the exemplars. In this work, we are not aiming for perfect imitations. We seek to create \emph{novel sounds} by via artifical, exploratory creativity. Boden defines this concept as an emergent property of generative work within confined rule sets~\cite{boden2009computer}. Ultimately, we wish to work within the limitations of any tractable sound source to create its approximations of a given sound category. Within the music realm, a prominent instance of exploratory creativity is the persistent aesthetics of 8-bit soundtracks, remaining popular long after the original constraints are lifted~\cite{collins2007loop}.
    \item \textit{Interpretability}: Neural networks are often described as black boxes~\cite{basheer2000artificial}. Their highly recursive structure makes modern explanation methods such as saliency maps unreliable~\cite{rudin2019stop}.  
    \item \textit{Speed of Rendering at high sampling rates}: Despite the utilization of powerful GPUs, the standard sampling rate in most audio generation work utilizing neural is under 24 khz \cite{yamamoto2020parallel,oord2017parallel,aouameur2019neural,ramires2020neural}. However, a significant number of untrained human ears can detect a change in quality of audio between sampling rates of 192 khz and the industry standard of 44.1 khz \cite{reiss2016meta}. In this work, we fix our sampling rate 48 khz. 
\end{itemize}

 Evaluation of a periodic function such as sine or cosine is the simplest form of software audio generation~\cite{mitchell2009basicsynthChap5}. This generative system is called an \textit{oscillator}. With careful programming, DSP techniques can be used to replicate almost any sound, which is why they have powered commercial digital synthesizers for over half a century ~\cite{jenkins2019analog}. Synthesizers we build using these functions are tractable and explainable: the output is determined by the input and reproducible. This makes the evaluation of a set of inputs (or parameters) to our synthesizers relatively simple.

We opted for a set of classical DSP methods to build our synthesizer, which allow for fast offline generation of audio signals. We made extensive use of Pippi\footnote{https://github.com/luvsound/pippi} and SciPy~\cite{jones2001scipy} libraries. Our virtual synthesizer contains a set of one or more submodules. Each submodule is a self-contained noise making unit. Submodules have identical sets of parameters, but widely different outputs can be achieved depending on the values assigned. The set of parameters available to each submodule is highlighted in Table~\ref{table:submodule_params}. The sonic output of the virtual synthesizer is the normalized addition of the sonic output of its submodules. Our implementation of a synthesizer can have any number of submodules. The parameters that dictate the output signal of each submodule as well as the range of values each parameter can take are shown in table~\ref{table:submodule_params}. We call the number of submodules in each virtual synthesizer the \textit{stack size}. We call the sets of parameter values that characterize a synthesizer's submodules a \textit{program} (analogous to a preset for a VST).

@@

\begin{figure}[htbp]
    \begin{center}
    \textbf{Data Overview}
    \makebox[\textwidth]{
    \includegraphics[width=0.95\linewidth]{images/venn_data.pdf}}
    \end{center}
    \caption{ 
    Scales are symbolic (true scales are not measurable and/or subjective).  $\mathcal{N}$ is the set of percussive sounds a synthesizer is capable of making. Our positive samples, $\mathcal{T^{+}}$, is a small fraction of a wide variety of percussive sounds that are conceivable. The size of $\mathcal{T^{-}}$ can be set to any number(we can generate as any number of random samples), but we prefer to keep it no larger than  $\mathcal{T^{+}}$. $\mathcal{H}$ is a series of sounds sent to the ear for classification. The ideal ear will detect  $\mathcal{H^{+}}$, the overlap of $\mathcal{H}$ and $\mathcal{N}$, with high precision and recall. 
    }
\label{fig:ven_data}
\end{figure}
\hypref{Decision.1} requires knowledge of what drums \textbf{do not} sound like, or knowledge of an infinitely large set, which cannot be fully represented via examples. An important consideration is that the source of sounds used in training the model (organic drum sounds) will be fundamentally different from the source of unlabeled sounds we wish to categorize (noise from a synthesizer). This issue is reflective of the OSR problem~\cite{geng2020recent,mundt2019open}. Figure~\ref{fig:ven_data} highlights a number of caveats with this approach. we denote $\mathcal{N}$ as the set of percussive sounds a synthesizer is capable of making. $\mathcal{T^{+}}$, our positive examples, is a set of sound extracted from material drums. $\mathcal{T^{-}}$, is a small subset of an infinite set which is meant to represent all noises which the synthesizer is capable of making and likely includes sounds in $\mathcal{N}$. During the hearing test, the virtual ear continually receives synthesizer noises for classification. The set of received noises is denoted as $\mathcal{H}$. The performance of the ear is measured by its precision and recall in finding the overlap between $\mathcal{H}$ and $\mathcal{N}$. It is important to ensure that the change in learning domains ---particularly with the case of positive examples of percussive sounds--- does not interfere with transformation of knowledge from the training of the ear to the hearing test. To maximize the transference of knowledge from one domain to another, we seek to learn from agnostic feature sets that capture fundamental characteristics of the data points.
\subsection{Feature Extraction}
To train our virtual ear, we require generalizable, domain agnostic features. Various works have demonstrated effective reconstruction of signals given their Short-time Fourier Transforms (STFT) ~\cite{nawab1983signal,griffin1984signal}. If the STFT of a signal can be used its reconstruction, perhaps it can be utilized as a source of fundamental features necessary for audible signal categorization. We defined 3 STFT transformation functions which we believe to capture important, unique attributes of percussive sounds (figure ~\ref{fig:stackspectrums}). Our transformations are GPU based, allowing their application in tandem to the training of the classifier or tuning of STFT transformation during hyper-paramter optimization (see Appendix~\ref{appendix:hyperparam}).
\subsection{Ear Types}
\label{classifiers}
